---
title: "Debugging R Code"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "Help! What's wrong with my code???"
date: 2022-09-29
categories: [module 4, week 5, programming, debugging]
---

<!-- Add interesting quote -->

# Pre-lecture materials

### Read ahead

::: callout-note
## Read ahead

**Before class, you can prepare by reading the following materials:**

1. <https://adv-r.hadley.nz/debugging>
2. <https://rstats.wtf/debugging-r-code>

:::

### Acknowledgements

Material for this lecture was borrowed and adopted from

- <https://rdpeng.github.io/Biostat776/lecture-debugging-r-code>
- <https://adv-r.hadley.nz/debugging>

# Learning objectives

::: callout-note
# Learning objectives

**At the end of this lesson you will:**

- Discuss an overall approach to debugging code in R 
- Recognize the three main indications of a problem/condition (`message`, `warning`, `error`) and a fatal problem (`error`)
- Understand the importance of reproducing the problem when debugging a function or piece of code
- Learn how to use interactive debugging tools `traceback`, `debug`, `recover`, `browser`, and `trace` can be used to find problematic code in functions

:::

# Debugging R Code

## Overall approach

Finding the **root cause of a problem is always challenging**. 

Most bugs are subtle and hard to find because if they were obvious, you would have avoided them in the first place. 

A good strategy helps. Below I outline a four step process that I have found useful:

### 1. Google!

Whenever you see an error message, **start by googling it**. 

If you are lucky, you will discover that it’s a common error with a known solution.

:::{.callout-tip}

### Pro-tip

When googling, improve your chances of a good match by removing any variable names or values that are specific to your problem.

:::

### 2. Make it repeatable

To find the root cause of an error, you are going to need to execute the code many times as you consider and reject hypotheses. 

**To make that iteration as quick possible**, it’s worth some upfront investment to **make the problem both easy and fast to reproduce**.

Start by creating a **rep**roducible **ex**ample (reprex). 

- This will help others help you, and **often leads to a solution without asking others**, because in the course of making the problem reproducible you often figure out the root cause. 

Make the **example minimal by removing code and simplifying data**. 

- As you do this, you may discover inputs that do not trigger the error. - Make note of them: they will be helpful when diagnosing the root cause.

:::{.callout-tip}

### Example

Let's try making a **reprex** [using the `reprex` package](https://www.tidyverse.org/help) (installed with the `tidyverse`)

```{r}
library(reprex)
```

Write a bit of code and copy it to the clipboard:

```{r}
#| eval: false
(y <- 1:4)
mean(y)
```

Enter `reprex()` in the R Console. In RStudio, you’ll see a preview of your rendered reprex.

It is now ready and waiting on your clipboard, so you can paste it into, say, a GitHub issue.

One last step. Let's go here and open up an issue on the course website: 

- <https://github.com/stephaniehicks/jhustatcomputing2022/issues>

We will paste in the code from our reprex.

:::

In RStudio, you can access reprex from the addins menu, which makes it even easier to point out your code and select the output format.


### 3. Figure out where it is

It’s a great idea to adopt the scientific method here. 

- Generate hypotheses
- Design experiments to test them
- Record your results

This may seem like a lot of work, but **a systematic approach** will end up saving you time. 

Often **a lot of time can be wasted relying on my intuition to solve a bug** (“oh, it must be an off-by-one error, so I’ll just subtract 1 here”), when I would have been better off taking a systematic approach.

If this fails, you **might need to ask help from someone else**. 

If you have followed the previous step, you will have a small example that is easy to share with others. That makes it much easier for other people to look at the problem, and more likely to help you find a solution.

### 4. Fix it and test it

Once you have found the bug, you need to **figure out how to fix it** and to **check that the fix actually worked**. 

Again, it is very useful to have automated tests in place. 

- Not only does this help to ensure that you **have actually fixed the bug**, it also **helps to ensure you have not introduced any new bugs** in the process. 
- In the absence of automated tests, make sure to **carefully record the correct output**, and check against the inputs that previously failed.


## Something’s Wrong!

Once you have made the error repeatable, the next step is to figure out where it comes from.

R has a number of **ways to indicate to you that something is not right**. 

There are **different levels of indication** that can be used, ranging from mere notification to fatal error. Executing any function in R may result in the following **conditions**.

- `message`: A **generic notification/diagnostic message** produced by the `message()` function; execution of the function continues
- `warning`: An indication that **something is wrong but not necessarily fatal**; execution of the function continues. Warnings are generated by the `warning()` function
- `error`: An indication that **a fatal problem has occurred** and execution of the function stops. Errors are produced by the `stop()` function.
- `condition`: A generic concept for indicating that **something unexpected has occurred**; programmers can create their own custom conditions if they want.

:::{.callout-tip}

### Example

Here is an example of a warning that you might receive in the course of using R.

```{r}
#| warning: true
log(-1)
```

This warning lets you know that taking the log of a negative number results in a `NaN` value because you **can't take the log of negative numbers**. 

:::

Nevertheless, R doesn't give an error, because it has a useful value that it can return, the **`NaN` value**. 

The **warning is just there** to let you know that **something unexpected happen**. 

Depending on what you are programming, you may have intentionally taken the log of a negative number in order to move on to another section of code.

:::{.callout-tip}

### Example

Here is another function that is designed to print a message to the console depending on the nature of its input.

```{r}
print_message <- function(x) {
        if(x > 0) {
                print("x is greater than zero")
        } else {
                print("x is less than or equal to zero")
        }  
        invisible(x)        
}
```

This function is simple: 

- It **prints a message** telling you whether `x` is greater than zero or less than or equal to zero.
- It also returns its input **invisibly**, which is a common practice with "print" functions. 

**Returning an object invisibly** means that the **return value does not get auto-printed** when the function is called.

Take a hard look at the function above and see if you can identify any bugs or problems.

We can execute the function as follows.

```{r}
#| error: true
print_message(1)
```

The function seems to work fine at this point. No errors, warnings, or messages.

```{r}
#| error: true
print_message(NA)
```

:::

What happened?

- Well, the first thing the function does is test if `x > 0`. 
- But you can't do that test if `x` is a `NA` or `NaN` value. 
- R **doesn't know what to do in this case** so it **stops with a fatal error**.

We can **fix this problem** by anticipating the possibility of `NA` values and checking to see if the input is `NA` with the `is.na()` function.

```{r}
print_message2 <- function(x) {
        if(is.na(x))
                print("x is a missing value!")
        else if(x > 0)
                print("x is greater than zero")
        else
                print("x is less than or equal to zero")
        invisible(x)
}
```

Now we can run the following.

```{r}
print_message2(NA)
```

And all is fine.

Now what about the following situation.

```{r}
#| error: true
x <- log(c(-1, 2))
print_message2(x)
```

Now what?? Why are we getting this warning? 

The **warning** says "the condition has length > 1 and only the first element will be used". 

The **problem here** is that I passed `print_message2()` a vector `x` that was of length 2 rather then length 1. 

Inside the body of `print_message2()` the expression `is.na(x)` returns a vector that is tested in the `if` statement. 

However, `if` cannot take vector arguments, so you get a warning. 

The fundamental problem here is that `print_message2()` is not **vectorized**.

We can **solve this problem** two ways. 

1. Simply **not allow vector arguments**. 
2. The other way is to **vectorize** the `print_message2()` function to allow it to take vector arguments.

For the **first way**, we simply need to check the length of the input.

```{r}
print_message3 <- function(x) {
        if(length(x) > 1L)
                stop("'x' has length > 1")
        if(is.na(x))
                print("x is a missing value!")
        else if(x > 0)
                print("x is greater than zero")
        else
                print("x is less than or equal to zero")
        invisible(x)
}
```

Now when we pass `print_message3()` a vector, we should get an **error**.

```{r}
#| error: true
print_message3(1:2)
```

Vectorizing the function can be accomplished easily with the `Vectorize()` function.

```{r}
print_message4 <- Vectorize(print_message2)
out <- print_message4(c(-1, 2))
```

You can see now that the **correct messages are printed without any warning or error**. 

:::{.callout-tip}

### Note

I stored the return value of `print_message4()` in a separate R object called `out`. 

This is because when I use the `Vectorize()` function it no longer preserves the invisibility of the return value.

:::

:::{.callout-tip}

### Helpful tips

The **primary task of debugging** any R code is **correctly diagnosing what the problem is**. 

When diagnosing a problem with your code (or somebody else's), it's important **first understand what you were expecting to occur**. 

Then you need to **idenfity what did occur** and **how did it deviate from your expectations**. 

Some basic questions you need to ask are

- What was your input? How did you call the function?
- What were you expecting? Output, messages, other results? 
- What did you get?
- How does what you get differ from what you were expecting? 
- Were your expectations correct in the first place?
- Can you reproduce the problem (exactly)?

:::

Being able to answer these questions is important not just for your own sake, but in situations where you may need to ask someone else for help with debugging the problem. 

Seasoned programmers will be asking you these exact questions.



# Debugging Tools in R

R provides a number of tools to help you with debugging your code. The primary tools for debugging functions in R are

- `traceback()`: **prints out the function call stack** after an error occurs; does nothing if there’s no error
- `debug()`: **flags a function for “debug” mode** which allows you to step through execution of a function one line at a time
- `browser()`: **suspends the execution of a function** wherever it is called and puts the function in debug mode
- `trace()`: allows you to **insert debugging code into a function** at specific places 
- `recover()`: allows you to **modify the error behavior** so that you can browse the function call stack

These functions are interactive tools specifically designed to allow you to pick through a function. There is also the more blunt technique of inserting `print()` or `cat()` statements in the function.

## Using `traceback()`

The `traceback()` function **prints out the function call stack** after an error has occurred. 

The **function call stack** is the **sequence of functions that was called before the error occurred**. 

For example, you may have a function `a()` which subsequently calls function `b()` which calls `c()` and then `d()`. 

If an error occurs, it may not be immediately clear in which function the error occurred.

The `traceback()` function **shows you how many levels deep** you were when the error occurred.

:::{.callout-tip}

### Example

Let's use the `mean()` function on a vector `z` that does not exist in our R environment

```r
> mean(z)
Error in mean(z) : object 'z' not found
> traceback()
1: mean(z)
```

Here, it's **clear that the error occurred** inside the `mean()` function because the object `z` does not exist. 

:::

The `traceback()` function **must be called immediately after an error** occurs. Once another function is called, you lose the traceback.


:::{.callout-tip}

### Example

Here is a slightly more complicated example using the `lm()` function for linear modeling.

```r
> lm(y ~ x)
Error in eval(expr, envir, enclos) : object ’y’ not found
> traceback()
7: eval(expr, envir, enclos)
6: eval(predvars, data, env)
5: model.frame.default(formula = y ~ x, drop.unused.levels = TRUE)
4: model.frame(formula = y ~ x, drop.unused.levels = TRUE)
3: eval(expr, envir, enclos)
2: eval(mf, parent.frame())
1: lm(y ~ x)
```

You can see now that the **error did not get thrown until the 7th level of the function call stack**, in which case the `eval()` function tried to evaluate the formula `y ~ x` and **realized the object `y` did not exist**.

:::

Looking at the traceback is useful for figuring out roughly where an error occurred but it's not useful for more detailed debugging. For that you might turn to the `debug()` function.

## Using `debug()`

<details> <summary> Click here for how to use `debug()` with an interactive browser.</summary>

The `debug()` function initiates an interactive debugger (also known as the "browser" in R) for a function. With the debugger, you can step through an R function one expression at a time to pinpoint exactly where an error occurs.

The `debug()` function takes a function as its first argument. Here is an example of debugging the `lm()` function. 

```r
> debug(lm)      ## Flag the 'lm()' function for interactive debugging
> lm(y ~ x)
debugging in: lm(y ~ x)
debug: {
    ret.x <- x
    ret.y <- y
    cl <- match.call()
    ...
    if (!qr)
        z$qr <- NULL 
    z
} 
Browse[2]>
```

Now, every time you call the `lm()` function it will launch the interactive debugger. To turn this behavior off you need to call the `undebug()` function.

The debugger calls the browser at the very top level of the function body. From there you can step through each expression in the body. There are a few special commands you can call in the browser:

* `n` executes the current expression and moves to the next expression
* `c` continues execution of the function and does not stop until either an error or the function exits
* `Q` quits the browser

Here's an example of a browser session with the `lm()` function.

```r
Browse[2]> n   ## Evalute this expression and move to the next one
debug: ret.x <- x
Browse[2]> n
debug: ret.y <- y
Browse[2]> n
debug: cl <- match.call()
Browse[2]> n
debug: mf <- match.call(expand.dots = FALSE)
Browse[2]> n
debug: m <- match(c("formula", "data", "subset", "weights", "na.action",
    "offset"), names(mf), 0L)
```

While you are in the browser you can execute any other R function that might be available to you in a regular session. In particular, you can use `ls()` to see what is in your current environment (the function environment) and `print()` to print out the values of R objects in the function environment.

You can turn off interactive debugging with the `undebug()` function.

```r
undebug(lm)    ## Unflag the 'lm()' function for debugging
```

</details>


## Using `recover()`

<details> <summary> Click here for how to use `recover()` with an interactive browser.</summary>

The `recover()` function can be used to modify the error behavior of R when an error occurs. Normally, when an error occurs in a function, R will print out an error message, exit out of the function, and return you to your workspace to await further commands. 

With `recover()` you can tell R that when an error occurs, it should halt execution at the exact point at which the error occurred. That can give you the opportunity to poke around in the environment in which the error occurred. This can be useful to see if there are any R objects or data that have been corrupted or mistakenly modified.

```r
> options(error = recover)    ## Change default R error behavior
> read.csv("nosuchfile")      ## This code doesn't work
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file ’nosuchfile’: No such file or directory
  
Enter a frame number, or 0 to exit

1: read.csv("nosuchfile")
2: read.table(file = file, header = header, sep = sep, quote = quote, dec =
3: file(file, "rt")

Selection:
```

The `recover()` function will first print out the function call stack when an error occurrs. Then, you can choose to jump around the call stack and investigate the problem. When you choose a frame number, you will be put in the browser (just like the interactive debugger triggered with `debug()`) and will have the ability to poke around.

</details>

# Summary


- There are three main indications of a problem/condition: `message`, `warning`, `error`; only an `error` is fatal
- When analyzing a function with a problem, make sure you can reproduce the problem, clearly state your expectations and how the output differs from your expectation
- Interactive debugging tools `traceback`, `debug`, `recover`, `browser`, and `trace` can be used to find problematic code in functions
- Debugging tools are not a substitute for thinking!



# Post-lecture materials

### Final Questions

Here are some post-lecture questions to help you think about the material discussed.

::: callout-note
### Questions

1. Try using `traceback()` to debug this piece of code:

```{r, error=TRUE}
f <- function(a) g(a)
g <- function(b) h(b)
h <- function(c) i(c)
i <- function(d) {
  if (!is.numeric(d)) {
    stop("`d` must be numeric", call. = FALSE)
  }
  d + 10
}
f("a")
```

Describe in words what is happening above?

:::

### Additional Resources

::: callout-tip

- <https://rdpeng.github.io/Biostat776/lecture-debugging-r-code>
- <https://adv-r.hadley.nz/debugging>
- <https://rstats.wtf/debugging-r-code>

:::

