---
title: "Error Handling and Generation"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "Implement exception handling routines in R functions"
date: 2022-09-29
categories: [module 4, week 5, programming, debugging]
---

# Pre-lecture materials

### Read ahead

::: callout-note
## Read ahead

**Before class, you can prepare by reading the following materials:**

1. <https://adv-r.hadley.nz/debugging>
:::

### Acknowledgements

Material for this lecture was borrowed and adopted from

- <https://rdpeng.github.io/Biostat776/lecture-error-handling-and-generation>
- <https://adv-r.hadley.nz/debugging>


# Learning objectives

::: callout-note
# Learning objectives

**At the end of this lesson you will:**

- Create errors, warnings, and messages in R functions using the functions `stop`, `stopifnot`, `warning`, and `message`.
- Understand the importance of providing useful error messaging to improve user experience with functions. However, these can also slow down code substantially.

:::

# Error Handling and Generation

## What is an error?

Errors most often occur when code is used in a way that it is not intended to
be used. For example adding two strings together produces the following error:

```{r}
#| error: true
"hello" + "world"
```

The `+` operator is essentially a function that takes two numbers as arguments
and finds their sum. Since neither `"hello"` nor `"world"` are numbers, the
R interpreter produces an error. Errors will stop the execution of your program,
and they will (hopefully) print an error message to the R console.

In R there are two other constructs which are related to errors:
warnings and messages. Warnings are meant to indicate that something seems to
have gone wrong in your program that should be inspected. Here's a simple
example of a warning being generated:

```{r}
#| warning: true
as.numeric(c("5", "6", "seven"))
```

The `as.numeric()` function attempts to convert each string in 
`c("5", "6", "seven")` into a number, however it is impossible to convert
`"seven"`, so a warning is generated. Execution of the code is not halted,
and an `NA` is produced for `"seven"` instead of a number.

Messages simply print to the R console, though they are generated by an
underlying mechanism that is similar to how errors and warning are generated.
Here's a small function that will generate a message:

```{r}
#| message: true
f <- function(){
  message("This is a message.")
}

f()
```

## Generating Errors

There are a few essential functions for generating errors, warnings, and
messages in R. The `stop()` function will generate an error. Let's generate
an error:

```{r}
#| eval: false
stop("Something erroneous has occurred!")
```

```{r}
#| eval: false
Error: Something erroneous has occurred!
```

If an error occurs inside of a function then the name of that function will
appear in the error message:

```{r}
#| error: true
name_of_function <- function(){
  stop("Something bad happened.")
}

name_of_function()
```

The `stopifnot()` function takes a series of logical expressions as arguments
and if any of them are false an error is generated specifying which expression
is false. Let's take a look at an example:

```{r}
#| error: true
error_if_n_is_greater_than_zero <- function(n){
  stopifnot(n <= 0)
  n
}

error_if_n_is_greater_than_zero(5)
```

The `warning()` function creates a warning, and the function itself is very
similar to the `stop()` function. Remember that a warning does not stop the
execution of a program (unlike an error.)

```{r}
#| warning: true
warning("Consider yourself warned!")
```

Just like errors, a warning generated inside of a function will include the name
of the function in which it was generated:

```{r}
#| warning: true
make_NA <- function(x){
  warning("Generating an NA.")
  NA
}

make_NA("Sodium")
```

Messages are simpler than errors or warnings; they just print strings to the
R console. You can issue a message with the `message()` function:

```{r}
message("In a bottle.")
```

## When to generate errors or warnings

Stopping the execution of your program with `stop()` should only happen in the
event of a catastrophe - meaning only if it is impossible for your program to
continue. If there are conditions that you can anticipate that would cause
your program to create an error, then you should document those conditions so
whoever uses your software is aware. An example includes: 

- Providing invalid arguments to a function. You could check this at the beginning of your
program using `stopifnot()` so that the user can quickly realize something has gone wrong.

You can think of a function as kind of contract between you and the user: 

- if the user provides specified arguments, your program will provide predictable
results. Of course it's impossible for you to anticipate all of the potential
uses of your program. It's appropriate to create
a warning when this contract between you and the user is violated. 

A perfect example of this situation is the result of 

```{r}
as.numeric(c("5", "6", "seven"))
```

The user expects a vector of numbers to be returned as the
result of `as.numeric()` but `"seven"` is coerced into being NA, which is not
completely intuitive.

R has largely been developed according to the
[Unix Philosophy](https://en.wikipedia.org/wiki/Unix_philosophy), which generally
**discourages** printing text to the console unless something unexpected has
occurred. Languages that commonly run on Unix systems like C and C++ are
rarely used interactively, meaning that they usually underpin computer
infrastructure (computers "talking" to other computers). Messages printed to the
console are therefore not very useful since nobody will ever read them and it's
not straightforward for other programs to capture and interpret them.

In contrast, R code is frequently executed by human beings in the R console, which
serves as an interactive environment between the computer and person at the 
keyboard. If you think your program should produce a message, make sure that
the output of the message is primarily meant for a human to read. You should avoid
signaling a condition or the result of your program to another program by
creating a message.

## How should errors be handled?

Imagine writing a program that will take a long time to complete because of a
complex calculation or because you're handling a large amount of data. If an
error occurs during this computation then you're liable to lose all of the
results that were calculated before the error, or your program may not finish a
critical task that a program further down your pipeline is depending on. If you
anticipate the possibility of errors occurring during the execution of your
program, then you can design your program to handle them appropriately.

The `tryCatch()` function is the workhorse of handling errors and warnings in R.
The first argument of this function is any R expression, followed by conditions
which specify how to handle an error or a warning. The last argument, `finally`,
specifies a function or expression that will be executed after the expression
no matter what, even in the event of an error or a warning.

Let's construct a simple function I'm going to call 
[`beera`](https://en.wikipedia.org/wiki/Yogi_Berra) that catches errors and
warnings gracefully.

```{r}
beera <- function(expr){
  tryCatch(expr,
         error = function(e){
           message("An error occurred:\n", e)
         },
         warning = function(w){
           message("A warning occured:\n", w)
         },
         finally = {
           message("Finally done!")
         })
}
```

This function takes an expression as an argument and tries to evaluate it. If
the expression can be evaluated without any errors or warnings then the result
of the expression is returned and the message `Finally done!` is printed to the
R console. If an error or warning is generated, then the functions that are
provided to the `error` or `warning` arguments are printed. Let's try this
function out with a few examples.

```{r}
beera({
  2 + 2
})

beera({
  "two" + 2
})

beera({
  as.numeric(c(1, "two", 3))
})
```

Notice that we've effectively transformed errors and warnings into messages.

Now that you know the basics of generating and catching errors you'll need to
decide when your program should generate an error. My advice to you is to limit
the number of errors your program generates as much as possible. Even if you
design your program so that it's able to catch and handle errors, the error
handling process slows down your program by orders of magnitude. Imagine you
wanted to write a simple function that checks if an argument is an even number.
You might write the following:

```{r, error=TRUE}
is_even <- function(n){
  n %% 2 == 0
}

is_even(768)

is_even("two")
```

You can see that providing a string causes this function to raise an error. You
could imagine though that you want to use this function across a list of
different data types, and you only want to know which elements of that list are
even numbers. You might think to write the following:

```{r}
is_even_error <- function(n){
  tryCatch(n %% 2 == 0,
           error = function(e){
             FALSE
           })
}

is_even_error(714)

is_even_error("eight")
```

This appears to be working the way you intended, however when applied to more
data this function will be seriously slow compared to alternatives. For example
I could check that `n` is numeric before treating `n` like a number:

```{r}
is_even_check <- function(n){
  is.numeric(n) && n %% 2 == 0
}

is_even_check(1876)

is_even_check("twelve")
```

:::keyideas 

Notice that by using `is.numeric()` before the "AND" operator (`&&`), the
expression `n %% 2 == 0` is never evaluated. This is a programming language
design feature called "short circuiting." The expression can never evaluate to
`TRUE` if the left hand side of `&&` evaluates to `FALSE`, so the right hand
side is ignored.

:::

To demonstrate the difference in the speed of the code, we will use the
`microbenchmark` package to measure how long it takes for each function to be
applied to the same data.

```{r}
#| eval: false
library(microbenchmark)
microbenchmark(sapply(letters, is_even_check))
```

```
Unit: microseconds
                           expr    min      lq     mean  median      uq     max neval
 sapply(letters, is_even_check) 46.224 47.7975 61.43616 48.6445 58.4755 167.091   100
```

```{r, eval=FALSE}
microbenchmark(sapply(letters, is_even_error))
```

```
Unit: microseconds
                           expr     min       lq     mean   median       uq      max neval
 sapply(letters, is_even_error) 640.067 678.0285 906.3037 784.4315 1044.501 2308.931   100
```

The error catching approach is nearly 15 times slower!

Proper error handling is an essential tool for any software developer so that
you can design programs that are error tolerant. Creating clear and informative
error messages is essential for building quality software. 

:::{.callout-tip}

### Pro-tip

One closing tip I
recommend is to put documentation for your software online, including the
meaning of the errors that your software can potentially throw. Often a user's
first instinct when encountering an error is to search online for that error
message, which should lead them to your documentation!

:::

# Summary 

- Errors, warnings, and messages can be generated within R code using the functions `stop`, `stopifnot`, `warning`, and `message`.

- Catching errors, and providing useful error messaging, can improve user experience with functions but can also slow down code substantially.




# Post-lecture materials

### Additional Resources

::: callout-tip

- <https://rdpeng.github.io/Biostat776/lecture-error-handling-and-generation>
- <https://adv-r.hadley.nz/debugging>

:::