---
title: "Control Structures"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "Introduction to control the flow of execution of a series of R expressions"
date: 2022-09-20
categories: [module 4, week 4, R, programming]
---

<!-- Add interesting quote -->


# Pre-lecture materials

### Read ahead

::: callout-note
## Read ahead

**Before class, you can prepare by reading the following materials:**

1.  <https://rafalab.github.io/dsbook/programming-basics>
2.  <https://r4ds.had.co.nz/iteration>
:::

### Acknowledgements

Material for this lecture was borrowed and adopted from

-   <https://rdpeng.github.io/Biostat776/lecture-control-structures>
-   <https://r4ds.had.co.nz/iteration>

# Learning objectives

::: callout-note
# Learning objectives

**At the end of this lesson you will:**

-   Be able to use commonly used control structures including `if`, `while`, `repeat`, and `for`
-   Be able to skip an iteration of a loop using `next`
-   Be able to exit a loop immediately using `break`
:::

# Control Structures

**Control structures** in R allow you to **control the flow of execution of a series of R expressions**. 

Basically, control structures allow you to put some "logic" into your R code, rather than just always executing the same R code every time. 

Control structures **allow you to respond to inputs or to features of the data** and execute different R expressions accordingly.

Commonly used control structures are

-   `if` and `else`: testing a condition and acting on it

-   `for`: execute a loop a fixed number of times

-   `while`: execute a loop *while* a condition is true

-   `repeat`: execute an infinite loop (must `break` out of it to stop)

-   `break`: break the execution of a loop

-   `next`: skip an interation of a loop


:::{.callout-tip}

### Pro-tip

Most control structures are not used in interactive sessions, but rather when writing functions or longer expressions.

However, these constructs do not have to be used in functions and it's a good idea to become familiar with them before we delve into functions.

:::

## `if`-`else`

The `if`-`else` combination is probably the most commonly used control structure in R (or perhaps any language). This structure allows you to test a condition and act on it depending on whether it's true or false.

For starters, you can just use the `if` statement.

``` r
if(<condition>) {
        ## do something
} 
## Continue with rest of code
```

The above code does nothing if the condition is false. If you have an action you want to execute when the condition is false, then you need an `else` clause.

``` r
if(<condition>) {
        ## do something
} 
else {
        ## do something else
}
```

You can have a series of tests by following the initial `if` with any number of `else if`s.

``` r
if(<condition1>) {
        ## do something
} else if(<condition2>)  {
        ## do something different
} else {
        ## do something different
}
```

Here is an example of a valid if/else structure.

```{r}
## Generate a uniform random number
x <- runif(1, 0, 10)  

if(x > 3) {
        y <- 10
} else {
        y <- 0
}
```

The value of `y` is set depending on whether `x > 3` or not. This expression can also be written a different, but equivalent, way in R.

```{r}
y <- if(x > 3) {
        10
} else { 
        0
}
```

Neither way of writing this expression is more correct than the other. Which one you use will depend on your preference and perhaps those of the team you may be working with.

Of course, the `else` clause is not necessary. You could have a series of if clauses that always get executed if their respective conditions are true.

``` r
if(<condition1>) {

}

if(<condition2>) {

}
```


:::{.callout-note}

### Question

Let's use the `palmerpenguins` dataset and write a if-else statement that

1. Randomly samples a value from a standard normal distribution (**Hint**: check out the `rnorm()` function in base R).
2. If the value is larger than 0, use `dplyr` functions to keep only the `Chinstrap` penguins. 
3. Otherwise, keep only the `Gentoo` penguins. 
4. Re-run the code 10 times and look at output. 

```{r}
#| message: false
# try it yourself

library(tidyverse)
library(palmerpenguins)
penguins 

```

:::


## `for` Loops

**For loops** are pretty much the only looping construct that you will need in R. While you may occasionally find a need for other types of loops, in my experience doing data analysis, I've found very few situations where a for loop was not sufficient.

In R, for loops take an iterator variable and assign it successive values from a sequence or vector. 

For loops are most commonly used for **iterating over the elements of an object** (list, vector, etc.)

```{r}
for(i in 1:10) {
        print(i)
}
```

This loop takes the `i` variable and in each iteration of the loop gives it values 1, 2, 3, ..., 10, executes the code within the curly braces, and then the loop exits.

The following three loops all have the same behavior.

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:4) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

The `seq_along()` function is commonly used in conjunction with for loops in order to generate an integer sequence based on the length of an object (in this case, the object `x`).

```{r}
## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```

It is not necessary to use an index-type variable.

```{r}
for(letter in x) {
        print(letter)
}
```

For one line loops, the curly braces are not strictly necessary.

```{r}
for(i in 1:4) print(x[i])
```

However, I like to use curly braces even for one-line loops, because that way if you decide to expand the loop to multiple lines, you won't be burned because you forgot to add curly braces (and you *will* be burned by this).

## Nested `for` loops

`for` loops can be nested inside of each other.

``` r
x <- matrix(1:6, 2, 3)

for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i, j])
        }   
}
```

Nested loops are commonly needed for multidimensional or hierarchical data structures (e.g. matrices, lists). Be careful with nesting though. Nesting beyond 2 to 3 levels often makes it difficult to read/understand the code. If you find yourself in need of a large number of nested loops, you may want to break up the loops by using functions (discussed later).

## `while` Loops

While loops begin by testing a condition. If it is true, then they execute the loop body. Once the loop body is executed, the condition is tested again, and so forth, until the condition is false, after which the loop exits.

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```

While loops can potentially result in infinite loops if not written properly. Use with care!

Sometimes there will be more than one condition in the test.

```{r}
z <- 5
set.seed(1)

while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)
```

Conditions are always evaluated from left to right. For example, in the above code, if `z` were less than 3, the second test would not have been evaluated.

## `repeat` Loops

`repeat` initiates an infinite loop right from the start. These are not commonly used in statistical or data analysis applications but they do have their uses. The only way to exit a `repeat` loop is to call `break`.

One possible paradigm might be in an iterative algorithm where you may be searching for a solution and you don't want to stop until you're close enough to the solution. In this kind of situation, you often don't know in advance how many iterations it's going to take to get "close enough" to the solution.

```{r,eval=FALSE,prompt=FALSE}
x0 <- 1
tol <- 1e-8

repeat {
        x1 <- computeEstimate()
        
        if(abs(x1 - x0) < tol) {  ## Close enough?
                break
        } else {
                x0 <- x1
        } 
}
```

Note that the above code will not run if the `computeEstimate()` function is not defined (I just made it up for the purposes of this demonstration).

The loop above is a bit dangerous because there's no guarantee it will stop. You could get in a situation where the values of `x0` and `x1` oscillate back and forth and never converge. Better to set a hard limit on the number of iterations by using a `for` loop and then report whether convergence was achieved or not.

## `next`, `break`

`next` is used to skip an iteration of a loop.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```

`break` is used to exit a loop immediately, regardless of what iteration the loop may be on.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
      print(i)

      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```

# Summary

-   Control structures like `if`, `while`, and `for` allow you to control the flow of an R program
-   Infinite loops should generally be avoided, even if (you believe) they are theoretically correct.
-   Control structures mentioned here are primarily useful for writing programs; for command-line interactive work, the "apply" functions are more useful.

# Post-lecture materials

### Final Questions

Here are some post-lecture questions to help you think about the material discussed.

::: callout-note
### Questions

1.  Write for loops to compute the mean of every column in `mtcars`.

2.  Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.

3.  What happens if you use `for (nm in names(x))` and `x` has no names? What if only some of the elements are named? What if the names are not unique?
:::

### Additional Resources

::: callout-tip
-   <https://rdpeng.github.io/Biostat776/lecture-control-structures>
-   <https://rafalab.github.io/dsbook/programming-basics>
-   <https://r4ds.had.co.nz/iteration>
:::
